---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grates

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/reconverse/grates/branch/main/graph/badge.svg)](https://app.codecov.io/gh/reconverse/grates?branch=main)
[![R-CMD-check](https://github.com/reconverse/grates/workflows/R-CMD-check/badge.svg)](https://github.com/reconverse/grates/actions)
[![](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-maturing.svg)](https://www.reconverse.org/lifecycle.html#maturing)
[![R-CMD-check](https://github.com/reconverse/grates/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/reconverse/grates/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

grates provides a simple and coherent implementation of grouped date classes:

* *grates_yearweek* with arbitrary first day of the week (see `as_yearweek()`);
* *grates_month* (see `as_month()`);
* *grates_quarter* (see `as_quarter()`);
* *grates_year* (see `as_year()`);
* *grates_period* and *grates_int_period* for periods of constant length (see `as_period()` and `as_int_period()`).

These classes aim to be formalise the idea of a grouped date whilst also being intuitive in their use.
They build upon ideas of Davis Vaughan and the [datea](https://github.com/DavisVaughan/datea/) package.

For each of the grouped date classes, grates also provides scales to use with [ggplot2](https://cran.r-project.org/package=ggplot2). 

## Installation

The development version, which this documentation refers to, can be installed
from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!require(remotes)) {
  install.packages("remotes")
}
remotes::install_github("reconverse/grates", build_vignettes = TRUE)
```

You can install the current version of the package from either the releases [page](https://github.com/reconverse/grates/releases) or directly from [CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("grates")
```


## Vignette

A short illustration to grates functionality is provided in the worked example below but a more detailed introduction can be found in the included vignette

```r
`vignette("introduction", package = "grates")`
```

## Example
```{r, message=FALSE}
library(grates) 
library(outbreaks)  # for data
library(dplyr)      # for data manipulation
library(ggplot2)    # for plotting

# load some simulated linelist data
dat <- ebola_sim_clean$linelist

# group by week
weekly_dat <- 
  dat %>%
  mutate(date = as_yearweek(date_of_infection, firstday = 2)) %>% 
  count(date, name = "cases") %>% 
  na.omit()

head(weekly_dat, 8)

# plot
ggplot(weekly_dat, aes(date, cases)) + geom_col(width = 1, colour = "white") + theme_bw() + xlab("")
```

We make working with `<grates_yearweek>` and other grouped date objects easier by adopting logical conventions:

```{r yearweekconventions, error=TRUE}
dates <- as.Date("2021-01-01") + 0:30
weeks <- as_yearweek(dates, firstday = 5) # firstday = 5 to match first day of year
head(weeks, 8)
str(weeks)
dat <- tibble(dates, weeks)

# addition of wholenumbers will add the corresponding number of weeks to the object
mutate(dat, plus4 = weeks + 4)

# addition of two yearweek objects will error as it is unclear what the intention is
mutate(dat, addweeks = weeks + weeks)

# Subtraction of wholenumbers works similarly to addition
mutate(dat, minus4 = weeks - 4)

# Subtraction of two yearweek objects gives the difference in weeks between them
mutate(dat, plus4 = weeks + 4, difference = plus4 - weeks)

# weeks can be combined if they have the same firstday but not otherwise
wk1 <- as_yearweek("2020-01-01")
wk2 <- as_yearweek("2021-01-01")
c(wk1, wk2)
wk3 <- as_yearweek("2020-01-01", firstday = 2)
c(wk1, wk3)
```
