---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped date classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

grates tries to make it easy to perform common date grouping operation by providing a simple and coherent implementations of a variety ofgrouped date classes:

* *grates_yearweek* with arbitrary first day of the week (see `as_yearweek()`);
* *grates_month* (see `as_month()`);
* *grates_quarter* (see `as_quarter()`);
* *grates_year* (see `as_year()`);
* *grates_period* and *grates_int_period* for periods of constant length (see `as_period()` and `as_int_period()`).

These classes aim to be formalise the idea of a grouped date whilst also being intuitive in their use.
They build upon ideas of Davis Vaughan and the [datea](https://github.com/DavisVaughan/datea/) package.

As well as the class implementations we also provide x-axis scales for use with [ggplot2](https://cran.r-project.org/package=ggplot2). 

**NOTE**: When plotting graphs you may want more flexibility then the built in scales provide.
If you find yourself in this situation there is always the option to convert the grouped object to a date using the provided `as.Date` method.
This will return the date at the lower bound of the grouping. Equally you can convert to a factor with `as.factor`.

```{r setup, message=FALSE}
library(grates)
library(ggplot2)
library(outbreaks)
library(dplyr)
```

## grates_yearweek
`as_yearweek()` allows you to create *grates_yearweek* objects.
As arguments it takes, `x`, the date vector you wish to group and `firstday`, the day of the week you wish your weeks to start on;
(this defaults to 1 (Monday) and can go up to 7 (Sunday)).
The first week of the year is then defined as the first week containing 4 days in the new calendar year.
This means that the calendar year can sometimes be different to that of the associated `yearweek` object.

```{r yearweekdemo}
# create weekday names
wdays <- weekdays(as.Date(as_yearweek(as.Date("2020-01-01"), firstday = 1L)) + 0:6)
wdays <- setNames(1:7, wdays)

# example of how weeks vary by firstday over December and January
dates <- as.Date("2020-12-29") + 0:5
dat <- lapply(wdays, function(x) as_yearweek(dates, x))
bind_cols(dates = dates, dat)
```

We make working with `yearweek` and other grouped date objects easier by adopting logical conventions:

```{r yearweekconventions, error=TRUE}
dates <- as.Date("2021-01-01") + 0:30
weeks <- as_yearweek(dates, firstday = 5) # firstday = 5 to match first day of year
head(weeks, 8)
str(weeks)
dat <- tibble(dates, weeks)

# addition of wholenumbers will add the corresponding number of weeks to the object
dat %>% 
  mutate(plus4 = weeks + 4)

# addition of two yearweek objects will error as it is unclear what the intention is
dat %>% 
  mutate(plus4 = weeks + weeks)

# Subtraction of wholenumbers works similarly to addition
dat %>% 
  mutate(minus4 = weeks - 4)

# Subtraction of two yearweek objects gives the difference in weeks between them
dat %>% 
  mutate(plus4 = weeks + 4, difference = plus4 - weeks)

# weeks can be combined if they have the same firstday but not otherwise
wk1 <- as_yearweek("2020-01-01")
wk2 <- as_yearweek("2021-01-01")
c(wk1, wk2)
wk3 <- as_yearweek("2020-01-01", firstday = 2)
c(wk1, wk3)

# load some simulated linelist data
dat <- ebola_sim_clean$linelist

# Example of week plot
week_plot <- 
  dat %>%
  mutate(week = as_yearweek(date_of_infection), firstday = 7) %>% 
  count(week, name = "cases") %>% 
  na.omit() %>% 
  ggplot(aes(week, cases)) + geom_col(width = 1, colour = "white") + theme_bw()

week_plot
```

We can have date labels on the x_axis by utilising `scale_x_grate_yearweek`:

```{r yearweekplots2}
week_plot + scale_x_grates_yearweek(format = "%Y-%m-%d", firstday = 7)
```

## grates_month
`as_month` allows users to group by a fixed number of months.
As arguments it takes: 

+ `x`, the date vector you wish to group;
+ `n`, the number of months you wish to group by (defaulting to 1) and; 
+ `origin`, an optional value indicating where you would like to start your periods from relative to the Unix Epoch (1970-01-01).

By default the provided scale creates a histogram-like plot (unlike a histogram the widths of the intervals in the plot are identical across months) but this can be changed to have centralised labelling.

```{r month}
month_dat <- 
  dat %>%
  mutate(date = as_month(date_of_infection, n = 2)) %>% 
  count(date, name = "cases") %>% 
  na.omit()

month_plot <- 
  ggplot(month_dat, aes(date, cases)) + 
    geom_col(width = 2, colour = "white") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    xlab("")

month_plot

month_plot + scale_x_grates_month(format = NULL, n = 2, origin = 0)
```

## grate_quarter and grate_year
`as_quarter()` and `as_year()` behave similarly to `as_yearweek()` with the main difference main being that they have no need for a `firstday` argument:

```{r yearothers}
# create weekday names
dates <- seq(from = as.Date("2020-01-01"), to = as.Date("2021-12-01"), by = "1 month")

as_quarter(dates)
as_year(dates)
as_quarter(dates[1]) + 0:1
as_year(dates[1]) + 0:1

dat %>%
  mutate(date = as_quarter(date_of_infection)) %>% 
  count(date, name = "cases") %>% 
  na.omit() %>% 
  ggplot(aes(date, cases)) + 
    geom_col(width = 1, colour = "white") + 
    scale_x_grates_quarter(n.breaks = 10) +
    theme_bw() + 
    xlab("")

dat %>%
  mutate(date = as_year(date_of_infection)) %>% 
  count(date, name = "cases") %>% 
  na.omit() %>% 
  ggplot(aes(date, cases)) + 
    geom_col(width = 1, colour = "white") + 
    scale_x_grates_year(n.breaks = 2) +
    theme_bw() + 
    xlab("")
```

## period
`as_period()` is similar to `as_month()` and allows users to group by periods of a fixed length.
As arguments it takes, `x`, the date or integer vector you wish to group, `n` and `origin`.

```{r period}
dat %>%
  mutate(date = as_period(date_of_infection, n = 14)) %>% 
  count(date, name = "cases") %>% 
  na.omit() %>% 
  ggplot(aes(date, cases)) + 
    geom_col(width = 14, colour = "white") + 
    theme_bw() + 
    xlab("")

dat %>%
  mutate(date = as_period(date_of_infection, n = 28)) %>% 
  count(date, name = "cases") %>% 
  na.omit() %>% 
  ggplot(aes(date, cases)) + 
    geom_col(width = 28, colour = "white") + 
    theme_bw() + 
    xlab("")
```
